# Reproducible Research: Peer Assessment 1

## Loading and preprocessing the data
```{r read in and process}
activityData <- read.csv(file="activity.csv") 
```

## What is mean total number of steps taken per day?
```{r get mean steps per day}
meanSteps <- aggregate(steps ~ date, activityData, mean)
medianSteps <- aggregate(steps ~ date, activityData, median)
hist(x=meanSteps$steps, main="Average steps per day", xlab="Mean Steps per day")
meanSteps
medianSteps
```
 
## What is the average daily activity pattern?
```{r step information per interval}
stepsPerInterval <- aggregate(steps ~ interval, activityData, mean)
plot(x=stepsPerInterval$interval, y=stepsPerInterval$steps, type="l", xlab = "Interval", ylab = "Avg qty of steps")
maxInterval <- head(stepsPerInterval[order(stepsPerInterval$steps, decreasing=TRUE),], n=1)
print(paste("Highest average steps per interval is for interval: " , maxInterval$interval, "with", maxInterval$steps))
```
## Imputing missing values
This will get us a count of how many intervals are missing step counts
```{r}
cleanActData <- activityData[!is.na(activityData$steps),]
rowsWithNaSteps <- nrow(cleanActData)
print(paste("We have", rowsWithNaSteps, "intervals with no step count reported"))
```

Now we'll take the average step count for any given interval and replace NA values with the mean step count for that interval from other days. 
```{r}
#merge the data with the averge step count for a given interval
activityDataWithAvg <- merge(activityData, stepsPerInterval, by.x="interval", by.y="interval")
# sort it back into day/interval order
activityDataWithAvg <- activityDataWithAvg[order(activityDataWithAvg$date,activityDataWithAvg$interval),]
# overlay steps data
activityDataWithAvg$steps.x[is.na(activityDataWithAvg$steps.x)] <- activityDataWithAvg$steps.y[is.na(activityDataWithAvg$steps.x)]
# rename columns
colnames(activityDataWithAvg) <- c("interval","steps","date","avgIntSteps")

```
#### Calculate cleaned up mean and medians
```{r}
meanCleanSteps <- aggregate(steps ~ date, activityDataWithAvg, mean)
medianCleanSteps <- aggregate(steps ~ date, activityDataWithAvg, median)
hist(x=meanCleanSteps$s, main="Average steps per day", xlab="Mean Steps per day")
meanCleanSteps
medianCleanSteps
```
### Comparison of inputing missing data
Here are the average steps per day compared side by side. And with a matching ylim to scale them properly.
```{r}
par(mfrow=c(1,2))
hist(x=meanSteps$steps, main="Average steps per day", xlab="Mean Steps per day with NAs",ylim=c(0,30))
hist(x=meanCleanSteps$s, main="Average steps per day", xlab="Mean Steps per day with NAs replaced",ylim=c(0,30))
```
Using the mean step quantity for a given interval to update other intervals increases the average steps per day. 

## Are there differences in activity patterns between weekdays and weekends?
Lets build two data frames. One for weekends and one for weekdays. 
```{r}
library(lattice)
activityDataWithAvg$weekDayOrEnd <- ifelse(weekdays(as.Date(activityDataWithAvg$date)) %in% c("Saturday","Sunday"),"WeekEnd","WeekDay")
weekDayOrEndAvg <- aggregate(steps ~ interval + weekDayOrEnd , activityDataWithAvg, mean)
xyplot(steps ~ interval |weekDayOrEnd, data = weekDayOrEndAvg, layout =c(1,2),type="l", ylab="Number of Steps", xlab="Interval")

```
